---
title: "DRY coding"
author: "Data Science in Biomedicine - Week 5"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```
## Introduction

We'll be using an in-built data set from the MASS package. But, because the MASS package can be a bit problematic to install, I've provided the data for you. It's in .rds format, so we load it in using the `read_rds()` function.

For more info about the dataset, head to https://stat.ethz.ch/R-manual/R-devel/library/MASS/html/survey.html

```{r}
library(tidyverse)
library(knitr)

survey <- read_rds("./data/survey.rds")
```

## Exercise 1
Explore the data - what variables are in there, what variable types, how many variables and observations?
```{r}
str(survey)
```


## Exercise 2
Round every numeric variable to 0 decimal places.

```{r}
survey_numeric_rounded <- survey %>%
  mutate(across(where(is.numeric), round, 0)) %>%
  drop_na()


```


## Exercise 3
Using the rounded dataframe you created in Exercise 2, calculate the mean of each numeric variable in the dataset. Use kable() to insert a nicely formatted table into your R Markdown document.

```{r}
numeric_means <- survey_numeric_rounded %>%
  summarise(across(where(is.numeric), mean))

kable(numeric_means)

```


## Exercise 4
Now make exactly the same table, using the pivot functions. Again, use kable() to insert a nicely formatted table into your R Markdown document.
```{r}
numeric_means_tidy <- numeric_means %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Mean")

kable(numeric_means_tidy)

```

## Exercise 5
Students had initially expressed their height in either imperial or metric units, but someone converted everything into centimeters. Create a new variable, Height_original, which will show height in centimeters for students who expressed their height in metric units, and height in inches for students who prefer imperial units. 

NOTE: This one is a bit of a challenge. You may want to check out the `case_when()` function.

```{r}
survey_numeric_heights <- survey_numeric_rounded %>%
  select(Height, M.I) %>%
  mutate(
    Height_original = case_when(
      M.I == "Imperial" ~ Height / 2.54, M.I == "Metric" ~ Height
    )
    )
```


## Exercise 6
Create a table that shows the mean and standard deviation of height - in centimeters for students who used metric units, and in inches for students who used imperial units. Use kable() to insert a nicely formatted table into your R Markdown document.

```{r}
height_summary <- survey_numeric_heights %>%
  filter(!is.na(Height), !is.na(M.I)) %>%
  group_by(M.I) %>%
  summarise(
    Mean_Height = mean(Height_original),
    SD_Height = sd(Height_original)
  )

kable(height_summary, col.names = c("Measurement Units", "Mean Height", "Standard Deviation"))
```


## Exercise 7
Create a contingency table to see the relationship between exercise and smoking. Pivot your table to make it easier to read, and then use kable() to insert a nicely formatted table into your R Markdown document.

```{r}
contingency_table <- survey_numeric_rounded %>%
  group_by(Smoke, Exer) %>%
  tally()
  pivot_wider(
    names_from = Smoke,
    values_from = n
  )
  
kable(contingency_table, 
      col.names = c("Smoker Status", "Exercise", "Count"),
      caption = "Relationship between Exercise and Smoking")
```


## Exercise 8
Create a grid of Pulse histograms - one histogram for each combination of Sex and Exercise.

```{r}
contingency_table_sex <- survey_numeric_rounded %>%
  group_by(Sex, Exer, Pulse) %>%
  tally()
  pivot_wider(
    names_from = Sex,
    values_from = n
  )
  
ggplot(contingency_table_sex, aes(x = Pulse)) +
  geom_histogram(binwidth = 5, fill = "skyblue", color = "black") +
  facet_grid(Sex ~ Exer) +
  labs(title = "Histograms of Pulse by Sex and Exercise",
       x = "Pulse",
       y = "Count") +
  theme_minimal()

```


## Exercise 9
Create a new variable, Hnd.diff, that shows the difference in span between the writing and non-writing hand. Create a set of boxplots of this variable, with a separate boxplot for those who write with their right hand, and another one for those who write with their left hand.

```{r}
survey_hand_diff <- survey_numeric_rounded %>% 
  select(Wr.Hnd,NW.Hnd,W.Hnd) %>%
  mutate(Hnd.diff = Wr.Hnd - NW.Hnd)

ggplot(survey_hand_diff, aes(x= W.Hnd, y=Hnd.diff)) +
  geom_boxplot() +
  labs(title = "Boxplots of Difference in span between the writing and non-writing hand ",
       x = "Dominant Hand",
       y = "Hand span difference") +
  theme_minimal()


```

